---
title: "p8105_hw2_zq2227"
author: "Zixuan Qiu zq2227"
output: github_document
date: "9/28/2023"
---

# Homework 1

## Problem 1
```{r}
library(tidyverse)
```

```{r}
polsmonth_data = read_csv(file = "./pols-month.csv")
```

```{r}
unemployment_data = read_csv(file = "./unemployment.csv")
```

```{r}
snp_data = read_csv(file = "./snp.csv")
```

### clean the data in pols-month.csv

```{r}
polsmonth_data =
  read_csv(file = "./pols-month.csv") |>
  janitor::clean_names()|>
  separate(mon, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|>
  mutate(month=as.character(month))|>
   mutate(president = if_else(prez_gop==1, "gop", "dem"))|>
  select(-prez_gop,-prez_dem,-day)|>
  relocate(year, month, president,everything())
```

```{r}
polsmonth_data
```


### clean the data in snp.csv 
```{r}
snp_data = 
  read_csv(file = "./snp.csv")|>
  janitor::clean_names()|>
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(year,year=as.numeric(year))|>
  mutate(year,year=ifelse(year>2024,year-100,year+0))|>
  mutate(year,year=as.character(year))|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|>
  mutate(month=as.character(month))|>
  arrange(year,month)|>
  select(-day)
```
```{r}
snp_data
```
###  tidy the unemployment data

```{r}
unemployment_data = 
  read_csv(file = "./unemployment.csv")|>
  janitor::clean_names()|>
  pivot_longer( 
    jan:dec,
    names_to = "month", 
    values_to = "unemploy")|>
    mutate(year=as.character(year))|>
    mutate(month=str_to_title(month))

```
```{r}
unemployment_data
```

### mergetrashdata 
```{r}
firstdata = 
  inner_join(polsmonth_data, snp_data)#merge snp and pols
  
```
```{r}
binddata=inner_join(firstdata, unemployment_data)|>
  relocate(year, month, president, close, unemploy,everything())#merge unemployment into the result.
binddata
```
### Description
```{r}
binddata=
  mutate(binddata,president,president=as.factor(president))|>
  mutate(year,year=as.numeric(year))

summary(binddata)
```
* The record of the data  from the 1950 to 2014. it includes 11 variables.
year and month: time of data. president: whether the president was republican  or democratic. 
close: the closing values of the S&P stock. 
unemployment:percentage of unemployment.
gov_gop:  number of republican governors.
sen_gop:  number of republican senators.
rep_gop:  number of republican representatives. 
gov_dem:  number of democratic governors. 
sen_dem:  number of democratic senators. 
rep_dem:  number of democratic representatives. 

## Problem 2
```{r}
library(readxl)
```

### Mr.trash wheel
```{r}
mrtrash_data = 
  read_excel("./202207 Trash Wheel Collection Data.xlsx", sheet = 1,skip=1, n_max = 547)|># omit non-data entries  and rows that do not include dumpster-specific data.
  janitor::clean_names()|>#clean up names of columns in data frames
  select(-x15,-x16,-month,-year)|> # clean columns containing notes and no data entries
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|># change the data format and  variable names
  mutate(homes_powered,homes_powered=weight_tons * (500 / 30))|>
  mutate(homes_powered,homes_powered=round(homes_powered))|> # calculate the homes_powered 
  mutate(Name= "Mrtrash")|> #use reasonable variable names
  relocate(Name,dumpster,year, month,day, weight_tons,homes_powered,everything())# change the location of variables 
```
```{r}
mrtrash_data
```
### professor trash wheel
```{r}
protrash_data=
  read_excel("./202207 Trash Wheel Collection Data.xlsx", sheet = 2,skip=1, n_max = 94)|> #omit non-data entries and rows that do not include dumpster-specific data
  janitor::clean_names()|>
  select(-month,-year)|>
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|> 
  mutate(homes_powered,homes_powered=weight_tons * (500 / 30))|>
  mutate(homes_powered,homes_powered=round(homes_powered))|> # calculate the homes_powered 
  mutate(Name= "Professor")|> # use reasonable variable names
  relocate(Name,dumpster,year, month,day, weight_tons,homes_powered,everything())
```
```{r}
protrash_data
```

### Gwynnda trash wheel
```{r}
gwytrash_data=
  read_excel("./202207 Trash Wheel Collection Data.xlsx", sheet = 4,skip=1, n_max = 107)|> #omit non-data entries and rows
  janitor::clean_names()|>
  select(-month,-year)|>
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|>
  mutate(homes_powered,homes_powered=weight_tons * (500 / 30))|>
  mutate(homes_powered,homes_powered=round(homes_powered))|> # calculate the homes_powered 
  mutate(Name= "Gwynnda")|> #change the variables name 
  relocate(Name,dumpster, year, month,day, weight_tons,homes_powered,everything())
```
```{r}
gwytrash_data
```

### Merge
```{r}
mrpromerge= full_join(mrtrash_data,protrash_data)# use full join # combine professor trash wheel  with mrtrash wheel data.join_by(Name, dumpster, year, month, day,weight_tons, homes_powered, volume_cubic_yards, plastic_bottles,polystyrene, cigarette_butts, glass_bottles, grocery_bags,chip_bags)`

trashmerge = full_join(mrpromerge,gwytrash_data)# combine the three data in one data frame.join_by(Name, dumpster, year, month, day,weight_tons, homes_powered, volume_cubic_yards, plastic_bottles,polystyrene, cigarette_butts)`
trashmerge
```
```{r}
write_csv(trashmerge,"./trashmergedata") #Exporting data as cvs. name is trashmergedata
```


###  Description
```{r}
protrash= filter(trashmerge,Name== "Professor")
proweightsum= sum(protrash$weight_tons)
proweightsum
```
#### The total weight of trash collected by Professor Trash Wheel is 190.12tons.
```{r}
gwycigjuly= filter(trashmerge,Name== "Gwynnda",year=="2021",month=="Jul")
cigsum=sum(gwycigjuly$cigarette_butts)
cigsum
```
#### The total number of cigarette butts collected by Gwynnda in July of 2021 are 16300.
```{r}
summary(trashmerge)
```
####  According to the data frame, we could knwo the average homes powered by three trash wheels is 50.23 and there are 747 days rcords for three trash wheels.
```{r}
totaltrashsum= sum(trashmerge$weight_tons)
totaltrashsum
```
#### The total trash collection by three wheels are 2249.26 tons.
```{r}
cigmax=filter(trashmerge,cigarette_butts==310000)
cigmax
```
#### In 09/02/2014 the Mrtrash collect the 31000 cigarette buttes that are maximum cigarette in the data. 
#### The trash has 8 category:plastic_bottles,polystyrene , cigarette_butts , glass_bottles , grocery_bags, chip_bags , sports_balls , plastic_bags.



## Problem 3 
### Baseline data
```{r}
baseline_data = 
  read_csv("./MCI_baseline.csv", skip = 1)|>
  janitor::clean_names()|> # clean column 
  filter(age_at_onset!= ".")|># remove participants who do not meet the criteria :no MCI at baseline
  mutate(age_at_onset,age_at_onset=as.numeric(age_at_onset))|>
  mutate(sex,sex=ifelse(sex==1,"Male","Female"))|> # change the sex index from "1,0" to character
  mutate(apoe4,apoe4=ifelse(apoe4==1,"Carrier","Noncarrier")) # change the apoe4  from "1,0" to character "carriwe, noncarrier.
```
```{r}
baseline_data
```
*  (a)
```{r}
rawbaseline_data=
  read_csv("./MCI_baseline.csv", skip = 1)|>
  janitor::clean_names()

rawbaseline_data
nrow(rawbaseline_data)# number of participants were recruited in raw raw baseline data 
```
483 participants were recruited in raw raw baseline data 
```{r}
summary(baseline_data)
nrow(baseline_data) # number of participants develop MCI
```
97 participants develop MCI in basline data

```{r}
mean(baseline_data$current_age) #average baseline age in cleaned baseline data. 
```
The average baseline age in cleaned baseline data is 65.6.
```{r}
womennumber=filter(baseline_data,sex=="Female")
nrow(womennumber)# number of total women number in post-processed baseline data. 
womencarrier=filter(baseline_data,sex=="Female" & apoe4=="Carrier")
nrow(womencarrier)# number of women is APOE4 carrier#
ratio=nrow(womencarrier)/nrow(womennumber)
ratio # proportion of women in the study are APOE4 carriers: number of women carrier/  total women number in post-processed baseline data. 
```
The proportion of women in the study are APOE4 carriers :0.652

### Biomarker data
```{r}
amyloid_data=
  read_csv("./mci_amyloid.csv", skip=1)|>#import
  janitor::clean_names()|>
  rename(id=study_id)# change variables names same as baseline data
```
```{r}
amyloid_data
```

### merge baseline and amyloid
```{r}
onlyinbaseline= anti_join(baseline_data,amyloid_data,by="id")
onlyinbaseline
```
There are 3 participants ID appear in only the baseline. ID is 14,49,268 
```{r}
onlyinamyloid=anti_join(amyloid_data,rawbaseline_data,by="id")
onlyinamyloid
```
There are 12 participants ID appear in only the amyloid datasets.

### Combine the demographic and biomarker datasets 
```{r}
admerge=
  inner_join(baseline_data,amyloid_data, by="id")|>#  keep the participants that appear in both  basline and amyloid datasets.
  drop_na()|> # clean the NA data.
  mutate(sex,sex=as.factor(sex))|> # change sex to factor 
  mutate(apoe4,apoe4=as.factor(apoe4))|># change     APOE4 to factor 
  mutate(baseline,baseline=as.numeric(baseline))|>
  mutate(time_2,time_2=as.numeric(time_2))|>
  mutate(time_4,time_4=as.numeric(time_4))|>
  mutate(time_6,time_6=as.numeric(time_6))|>
  mutate(time_8,time_8=as.numeric(time_8))#change the variables: β42/40 ratio from chr to num so that esier to observation.
```
```{r}
admerge
summary(admerge)
write_csv(admerge,"./ADdieasedata") #Exporting data as cvs. name is ADdieasemergedata

```

### Observation and discussion
```{r}
carrier=filter(admerge,apoe4=="Carrier")
noncarrier=filter(admerge,apoe4=="Noncarrier")
```
```{r}
carriermean=tibble(meanvalue=c(mean(carrier$baseline),mean(carrier$time_2),mean(carrier$time_4),mean(carrier$time_6),mean(carrier$time_8)),time=c("baseline","time_2","time_4","time_6","time_8"),type="carrier")
carriermean
noncarriermean=tibble(meanvalue=c(mean(noncarrier$baseline),mean(noncarrier$time_2),mean(noncarrier$time_4),mean(noncarrier$time_6),mean(noncarrier$time_8)),time=c("baseline","time_2","time_4","time_6","time_8"),type="noncarrier")
noncarriermean
amyloidmcimean=bind_rows(carriermean,noncarriermean)

apoe4compareplot=ggplot(amyloidmcimean, aes(x =time , y = meanvalue, color= type,group=type)) +geom_line()+geom_point()+labs(title = "β42/40 ratio between APOE4Carrier and APOE4Non-carrier in MCI patients", x = "Time", y = "Mean")+geom_text(aes(label = sprintf("%.3f", meanvalue)), vjust = -0.5)

apoe4compareplot
ggsave("apoe4 ratio comparison.png",plot=apoe4compareplot)
```
* According  to the plot,in the participants group who developed the MCI, the APOE4carrier has a lower β42/40 ratio than APOE4 non carrier.Since the APOE4 is significantly associated with a higher risk of developing Alzheimer’s disease and the  β 42/40 ratio linked to  clinical symptoms of Alzheimer’s disease.The MCI APOE4 carrier has a higher risk of developing Alzheimer’s disease accompanied by a lower β 42/40 ratio over time.
A low β 42/40 also indicate the risk of Alzheimer’s disease.As  MCI patients age, their  β 42/40 ratio levels gradually decline. 







