---
title: "p8105_hw2_zq2227"
author: "Zixuan Qiu zq2227"
output: github_document
date: "9/28/2023"
---

# Homework 1

## Problem 1
```{r}
library(tidyverse)
```

```{r}
polsmonth_data = read_csv(file = "./pols-month.csv")
```

```{r}
unemployment_data = read_csv(file = "./unemployment.csv")
```

```{r}
snp_data = read_csv(file = "./snp.csv")
```

### clean the data in pols-month.csv

```{r}
polsmonth_data =
  read_csv(file = "./pols-month.csv") |>
  janitor::clean_names()|>
  separate(mon, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|>
  mutate(month=as.character(month))|>
   mutate(president = if_else(prez_gop==1, "gop", "dem"))|>
  select(-prez_gop,-prez_dem,-day)|>
  relocate(year, month, president,everything())
```

```{r}
polsmonth_data
```


### clean the data in snp.csv 
```{r}
snp_data = 
  read_csv(file = "./snp.csv")|>
  janitor::clean_names()|>
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(year,year=as.numeric(year))|>
  mutate(year,year=ifelse(year>2024,year-100,year+0))|>
  mutate(year,year=as.character(year))|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|>
  mutate(month=as.character(month))|>
  arrange(year,month)|>
  select(-day)
```
```{r}
snp_data
```
###  tidy the unemployment data

```{r}
unemployment_data = 
  read_csv(file = "./unemployment.csv")|>
  janitor::clean_names()|>
  pivot_longer( 
    jan:dec,
    names_to = "month", 
    values_to = "unemploy")|>
    mutate(year=as.character(year))|>
    mutate(month=str_to_title(month))

```
```{r}
unemployment_data
```

### mergetrashdata 
```{r}
firstdata = 
  inner_join(polsmonth_data, snp_data)#merge snp and pols
  
```
```{r}
binddata=inner_join(firstdata, unemployment_data)|>
  relocate(year, month, president, close, unemploy,everything())#merge unemployment into the result.
binddata
```
### Description
```{r}
binddata=
  mutate(binddata,president,president=as.factor(president))|>
  mutate(year,year=as.numeric(year))

summary(binddata)
```
* The record of the data  from the 1950 to 2014. it includes 11 variables.
year and month: time of data. president: whether the president was republican  or democratic. 
close: the closing values of the S&P stock. 
unemployment:percentage of unemployment.
gov_gop:  number of republican governors.
sen_gop:  number of republican senators.
rep_gop:  number of republican representatives. 
gov_dem:  number of democratic governors. 
sen_dem:  number of democratic senators. 
rep_dem:  number of democratic representatives. 

## Problem 2
```{r}
library(readxl)
```

### Mr.trash wheel
```{r}
mrtrash_data = 
  read_excel("./202309 Trash Wheel Collection Data.xlsx", sheet = 1,skip=1, n_max = 584)|># omit non-data entries  and rows that do not include dumpster-specific data.
  janitor::clean_names()|>#clean up names of columns in data frames
  select(-x15,-x16,-month,-year)|> # clean columns containing notes and no data entries
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|># change the data format and  variable names
  mutate(homes_powered,homes_powered=weight_tons * (500 / 30))|>
  mutate(homes_powered,homes_powered=round(homes_powered))|> # calculate the homes_powered 
  mutate(Name= "Mrtrash")|> #use reasonable variable names
  relocate(Name,dumpster,year, month,day, weight_tons,homes_powered,everything())# change the location of variables 
```
```{r}
mrtrash_data
```
### professor trash wheel
```{r}
protrash_data=
  read_excel("./202309 Trash Wheel Collection Data.xlsx", sheet = 2,skip=1, n_max = 106)|> #omit non-data entries and rows that do not include dumpster-specific data
  janitor::clean_names()|>
  select(-month,-year)|>
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|> 
  mutate(homes_powered,homes_powered=weight_tons * (500 / 30))|>
  mutate(homes_powered,homes_powered=round(homes_powered))|> # calculate the homes_powered 
  mutate(Name= "Professor")|> # use reasonable variable names
  relocate(Name,dumpster,year, month,day, weight_tons,homes_powered,everything())
```
```{r}
protrash_data
```

### Gwynnda trash wheel
```{r}
gwytrash_data=
  read_excel("./202309 Trash Wheel Collection Data.xlsx", sheet = 4,skip=1, n_max = 155)|> #omit non-data entries and rows
  janitor::clean_names()|>
  select(-month,-year)|>
  mutate(date=as.Date(date, format = "%m/%d/%y"))|>
  separate(date, into = c("year", "month","day"), sep = "-")|>
  mutate(month = month(as.numeric(month), label = TRUE, abbr = TRUE))|>
  mutate(homes_powered,homes_powered=weight_tons * (500 / 30))|>
  mutate(homes_powered,homes_powered=round(homes_powered))|> # calculate the homes_powered 
  mutate(Name= "Gwynnda")|> #change the variables name 
  relocate(Name,dumpster, year, month,day, weight_tons,homes_powered,everything())
```
```{r}
gwytrash_data
```

### Merge
```{r}
mrpromerge= full_join(mrtrash_data,protrash_data)# use full join # combine professor trash wheel  with mrtrash wheel data.join_by(Name, dumpster, year, month, day,weight_tons, homes_powered, volume_cubic_yards, plastic_bottles,polystyrene, cigarette_butts, glass_bottles, grocery_bags,chip_bags)`

trashmerge = full_join(mrpromerge,gwytrash_data)# combine the three data in one data frame.join_by(Name, dumpster, year, month, day,weight_tons, homes_powered, volume_cubic_yards, plastic_bottles,polystyrene, cigarette_butts)`
trashmerge
```
```{r}
write_csv(trashmerge,"./trashmergedata.csv") #Exporting data as cvs. name is trashmergedata
```


###  Description
```{r}
protrash= filter(trashmerge,Name== "Professor")
proweightsum= sum(protrash$weight_tons)
proweightsum
```
#### The total weight of trash collected by Professor Trash Wheel is 216.26tons.
```{r}
gwycigjuly= filter(trashmerge,Name== "Gwynnda",year=="2021",month=="Jul")
cigsum=sum(pull(gwycigjuly,cigarette_butts))
cigsum
```
#### The total number of cigarette butts collected by Gwynnda in July of 2021 are 16300.
```{r}
summary(trashmerge)
```
####  According to the data frame, we could knwo the average homes powered by three trash wheels is 50.21 and there are 845 days rcords for three trash wheels.
```{r}
totaltrashsum= sum(trashmerge$weight_tons)
totaltrashsum
```
#### The total trash collection by three wheels are 2543.01 tons.
```{r}
cigmax=filter(trashmerge,cigarette_butts==310000)
cigmax
```
#### In 09/03/2014 the Mrtrash collect the 31000 cigarette buttes that are maximum cigarette in the data. 
#### The trash has 7 category:plastic_bottles,polystyrene , cigarette_butts , glass_bottles , plastic_bags,  wrapper,sports ball.



## Problem 3 
### Baseline data
```{r}
baseline_data = 
  read_csv("./MCI_baseline.csv", skip = 1)|>
  janitor::clean_names()|> # clean column 
  mutate(age_at_onset=replace(age_at_onset, age_at_onset == ".", NA_character_))|>
  filter(current_age < age_at_onset|is.na(age_at_onset))|>#remove any participants who has MCI at baseline
  mutate(age_at_onset,age_at_onset=as.numeric(age_at_onset))|>
  mutate(sex,sex=ifelse(sex==1,"Male","Female"))|> # change the sex index from "1,0" to character
  mutate(apoe4,apoe4=ifelse(apoe4==1,"Carrier","Noncarrier")) # change the apoe4  from "1,0" to character "carrier, noncarrier.
```
```{r}
baseline_data
```

```{r}

nrow(baseline_data)# number of participants were recruited in  baseline data 
```
479 participants were recruited in baseline data 
```{r}
mcidevelop=filter(baseline_data, !is.na(age_at_onset))
nrow(mcidevelop) # number of participants develop MCI
```
93 participants develop MCI in basline data

```{r}
mean(pull(baseline_data,current_age)) #average baseline age in cleaned baseline data. 
```
The average baseline age in cleaned baseline data is 65,03.
```{r}
womennumber=filter(baseline_data,sex=="Female")
nrow(womennumber)# number of total women number in post-processed baseline data. 
womencarrier=filter(baseline_data,sex=="Female" & apoe4=="Carrier")
nrow(womencarrier)# number of women is APOE4 carrier#
ratio=nrow(womencarrier)/nrow(womennumber)
ratio # proportion of women in the study are APOE4 carriers: number of women carrier/  total women number in post-processed baseline data. 
```
The proportion of women in the study are APOE4 carriers :0.3

### Biomarker data
```{r}
amyloid_data=
  read_csv("./mci_amyloid.csv", skip=1)|>#import
  janitor::clean_names()|>
  rename(id=study_id)|> # change variables names same as baseline data
  pivot_longer(
    baseline:time_8,
    names_to = "time", 
    values_to = "ratio")# tidy the dataset of longitudinally observed biomarker values
```
```{r}
amyloid_data
```

### merge baseline and amyloid
```{r}
onlyinbaseline= anti_join(baseline_data,amyloid_data,by="id")
onlyinbaseline
```
There are 8 participants ID appear in only the baseline. 
```{r}
onlyinamyloid=anti_join(amyloid_data,baseline_data,by="id")
onlyinamyloid
nrow(onlyinamyloid)/5
```
There are 16 participants ID appear in only the amyloid datasets.

### Combine the demographic and biomarker datasets 
```{r}
admerge=
  inner_join(baseline_data,amyloid_data, by="id")|>#  keep the participants that appear in both  basline and amyloid datasets.
  mutate(sex,sex=as.factor(sex))|># change sex to factor 
  mutate(apoe4,apoe4=as.factor(apoe4))|># change APOE4 to factor 
  mutate(time,time=as.factor(time))|>
  mutate(ratio,ratio=as.numeric(ratio)) #change the variables: β42/40 ratio from chr to num so that esier to observation.
```
```{r}
admerge
write_csv(admerge,"./ADdieasedata.csv") #Exporting data as cvs. name is ADdieasemergedata

```

### Observation and discussion
```{r}
summary(admerge)
```
#### Description: Thee averge age of the merge data is 65.05 and there are 1025 female and 1330 male in the data. The average education years is 16.38 yesrs. In the datasets, there are 710 participants are apoe4 carrier. In the MCI patients, the averge age at the onset of MCI is  70.51 years old. The median valuse of the β42/40 ratio is 0.10967.
```{r}
apoe4compareplot=ggplot(admerge, aes(x = time, y = ratio)) +
  geom_boxplot() +
  facet_wrap(~ apoe4, ncol = 2) +
  labs(title = "β42/40 ratio between APOE4Carrier and APOE4Non-carrier", x = "Time", y = "β
 42/40 Ratio ")
apoe4compareplot
```
```{r}
ggsave("apoe4 ratio comparison.png",plot=apoe4compareplot)
```
* According  to the plot,in the baseline group, the APOE4carrier has a lower β42/40 ratio than APOE4 non carrier.Since the APOE4 is significantly associated with a higher risk of developing Alzheimer’s disease and the  β 42/40 ratio linked to  clinical symptoms of Alzheimer’s disease.The  APOE4 carrier has a higher risk of developing Alzheimer’s disease accompanied by a lower β 42/40 ratio over time.
A low β 42/40 also indicate the risk of Alzheimer’s disease.As  participants age, their  β 42/40 ratio levels gradually decline. 







